---
title: "FirstDraft"
author: "Yidie Feng & Qiaoyue Tang"
date: "3/20/2017"
output: md_document
---
```{r,warning=FALSE,message=FALSE,results='hide',include=FALSE}
library("ggplot2")
library(lme4)
library(reshape2)
library(dplyr)
```
#### Summary
This report examines the effectiveness of biosolids, a new kind of fertilizer that potentially enhances soil activity by providing food source through microbial activities. In particular, we compare the change in Mean Weight Diameter (MWD), a parameter measuring soil stability, and the change in cover value, a criterion for abundance of plant species between observations under biosolids and control. The impact of biosolids on soil stability is analyzed by a mixed effect model. The effect on abundance of plant species is addressed by linear regression. We also calculate the correlation between MWD and cover value to examine if they move in the same direction. We find strong evidence to support the hypothesis that biosolids significantly increases soil productivity and coverage of certain plant species.

#### 1. Introduction
Soil health and productivity is an important issue affecting our daily lives and social well-being. Our client, Miss Emma Avery, is completing her Master of Soil Science thesis which investigates the long term impact of biosolids application on grassland plant communities and soil health. Biosolids, the main interest of the experient, is a fertilizer of interest that provides food source for microbial activities thus enhancing soil productivity. The experiment was established in 2002 at OK Ranch, Jasmond, BC when four grasslands were randomly chosen and applied either biosolids or control (no biosolids). Later, the data were collected in 2016 and we received it in January 2017 for analysis. The primary interest is to investigate whether Mean Weight Diameter (MWD), a parameter that quantifies soil productivity, is affected by 1) treatment type and 2) sampling date. The secondary interest is to explore the long term effect of biosolids on specific plant species composition. The third interest is to examine the correlation between MWD and the cover value of plants. The results of this report assess the significance of biosolids application on soil productivity and plant composition. It potentially impacts the existing fertilizers and introduce a more effective kind of new product. Starting with a description of dataset and methods,the following report addresses the above questions of interests with detailed analysis and ends with conclusions and further discussions.

#### 2. Data Description
The experiment was laid out in 4 pieces of lands with similar characteristics which are treated as blocks. The investigator randomly applied biosolids to half of each block and no biosolids to the other half. Within each half of a block, 3 equally spaced transects were arranged. MWD data were obtained from 7 fixed sample spots along each transect. This process was conducted four times in April, June, August and October 2016 respectively. The dataset given to us includes the sampling month, block index, treatment type, transect number and the MWD averaged from those 7 soil sample for each transect.

Figure 2.1 shows the boxplots of MWD under both biosolids and control in 4 sampling dates. Figure 2.2 gives the similar idea but in terms of density. In April, August and October, we observe that data points under biosolids have higher MWD than those under control. It can be seen from the upward shift of data points under biosolids in the boxplots or from the rightward shift of density curves. In June, the data points under biosolids overlap with those under control, but the median under biosolids is higher (Figure 2.1). Similarly, in the density plot, even though there is no obvious evidence of a rightward shift of biosolids density, we see a larger proportion of observations with moderately high MWD values (between 1.5 to 2.0). Table 2.1 shows the mean MWD for 8 treatment-block combinations.

```{r,echo=FALSE}
soil<-read.table("MWD.csv", sep = ",", header = T) 
ggplot(aes(y=MWD, x=Treatment, fill = Treatment, alpha = 0.4), data=soil) + geom_boxplot() + geom_point() + facet_wrap(~Date) + labs(title="Boxplots of MWD over Sampling Dates")
```

Figure 2.1 Boxplots of MWD of biosolids and control group plotted by the four samping dates

```{r,echo=FALSE}
ggplot(aes(MWD, fill = Treatment, alpha = 0.4), data=soil) + geom_density() + facet_wrap(~Date) + labs(title="Density of MWD over Sampling Dates")
```

Figure 2.2 Density of MWD of biosolids and control group plotted by the four samping dates 

```{r,echo=FALSE}
group_by(soil,Date,Treatment) %>%
  summarise(mean(MWD), sd(MWD))
```

Table 2.1 Mean and standard deviation of MWD by sampling dates and treatment type 

For the investigation of plants, 5 plant transects were laid out in half of a block. The researchers randomly selected 10 plots from each transect. Visual assessment was done for each plot. The species present and the corresponding abundance was recorded. The measurement of plant composition was quantified by cover class, a number between 1 to 6 that describes the abundance of plant species. Cover class is then converted to cover value which is a number in percentage and is in one-to-one correspondence to cover class. Therefore there are 6 fixed levels of cover values in total. The dataset includes the plant species name, block index, treatment type, plot number, cover class and cover value. If researchers did not observe a species for certain plots, the entry of that species is omitted instead of being recorded as 0. In the following analysis, we specifically investigate one species, POPR.

Table 2.2 describes the frequencies of cover values appearing under biosolids application and control for all species. The bar graph (figure 2.3 and 2.4) provide a better visualization for the results in table 2.2. We observe that when changing from control group to biosolids group, there is a dramatic decrease in number of observations with low cover values (<=37.5), and an increase in number of observations with high cover values (>37.5). For the specific species POPR, table 2.3 summarizes the frequencies of observations under 6 levels of cover values for biosolids and control respectively. There is a significant difference between counts under two kinds of treatments, indicating that biosolids is potentially effective.

```{r,echo=FALSE}
pc <- read.csv("plant_cover.csv", header = T)
pc$Block <- as.factor(pc$Block)
group_by(pc,Block,Treatment) %>%
  summarise(mean(Cover.value), sd(Cover.value))
```

Table 2.1 Mean and standard deviation of cover value by block and treatment type 

```{r,echo=FALSE}
table_sum <- tapply(pc$Cover.value,pc$Treatment, table)
table_sum
```

Table 2.2 Number of observation in each class of cover value presented by treatment groups

```{r,echo=FALSE}
barplot(table_sum$Control, space = 0, ylim = c(0,700),xlab = "Cover Values", ylab = "Frenquency", main = "Counts of Cover Values with Control", col = "mistyrose")
```

Figure 2.3 Bar graph of frequencies in each class of cover value for control group

```{r,echo=FALSE}
barplot(table_sum$Biosolids, space = 0, ,ylim = c(0,700), xlab = "Cover Values", ylab = "Frenquency", main = "Counts of Cover Values with Biosolids", col = "mistyrose")
```

Figure 2.4 Bar graph of frequencies in each class of cover value for biosolids group

```{r,echo=FALSE}
pc.subset <- subset(pc, Species == "POPR")
table_sum_POPR <- tapply(pc.subset$Cover.value,pc.subset$Treatment, table)
table_sum_POPR
```

Table 2.3 Frequencies of observations in each class of cover value presented by treatment groups for species POPR

Figure 2.5 provides a visualization of change in cover value over different blocks under both biosolids and control for POPR. Each point represents the mean cover value under a specific treatment and block combination. The observed cover value under control group are close to 0 in all four blocks while the cover values under biosolids are quite high.

```{r,echo=FALSE}
pc.subset <- pc.subset[ ,c(3,4,9)]
by_blockTrt <- group_by(pc.subset, Block, Treatment)
dat.avg<-summarise(by_blockTrt, y.avg=sum(Cover.value)/50)
add2control <- c("2","Control",0)
dat.avg[8, ] <- add2control

dat.avg$y.avg <- as.numeric(dat.avg$y.avg)
ggplot(aes(x = Block, y = y.avg, group = Treatment, colour = Treatment), data = dat.avg) + geom_point() + geom_line() + labs(x = "Block", y = "Plant Cover Value", title = "Change in Cover Value of POPR over Different Blocks")
```

Figure 2.5 Change in cover value of species POPR over the four blocks for control and biosolids group

```{r,echo=FALSE}
ggplot(aes(y=Cover.value, x=Treatment, fill = Treatment, alpha = 0.4), data=pc.subset) + geom_boxplot() + geom_point() + facet_wrap(~Block)
```

Figure 2.6 Boxplots of cover value of species POPR in biosolids and control group plotted by the four blocks


#### 3. Methods
##### Long term impact of biosolids on soil (MWD)
We use mixed-effect linear model which has both fixed and random effects. In our context, MWD is the response variable. Treatment is a fixed effect because the experimenters directly manipulate the application of treatment to the soil. Date is also a fixed effect because the four sampling dates are predetermined by the researchers. Block is treated as a random effect because we use random effect to handle the issue of dependencies that arise from the experimental design. 

“Repeated measurements” is the case where observations are taken from the same subject several times. In our case, MWD is measured in roughly the same place four times during the growing season. Therefore, the four observations are dependent because they are affected by common characteristics belonging to the same sampling location. 

Another complication is the transects setup. It results in the problem of “Pseudoreplication” where replicated observations are not independent. There are two sources of pseudo replication in our case:  
* The seven soil samples along one transect are pseudo replicates because the soil composition measurement in one place will be highly correlated with that five feet away.   
* In every half of the block, the three transects are pseudo replicates because they belong to the same experimental unit. The conditions affecting one transect will also affect another transect in that half of the block. 

Therefore, both repeated measurements and pseudoreplication violate the assumption of independence that is assumed in linear model. The way we tackle this problem is to introduce a random block effect into the model. This mixed-effect linear model will then analyze the data in a way that is similar to randomized complete block design but with a random block effect. Although transect is not included in the model, it is not of particular interest to the researchers. Moreover, it is inappropriate to investigate the variation across transects since they are not entirely independent. 

##### Long term impact of biosolids on plant cover
The same problem of pseudoreplication also appears in the plant cover dataset. The observations within and between 5 plant transects are dependent for the same reasons described in soil transects. The method we use is to take the average of cover value over each treatment-block combination. Then for treatment group, we have four independent observations from the four blocks. For control group, we also have four independent observations from the four blocks. For a block-treatment combination in which experimenters did not observe any presence of POPR, we add a cover value of 0 for that specific combination. Then we can fit an usual linear model to this averaged dataset with the averaged cover value as the response variable and treatment as the explanatory variable.

In addition, averaging observations from each block-treatment combination solves the problem that the response variable is discrete. Since we take an average over 50 discrete observations for each combination, the averaged cover value would be more continuous. Therefore the usual linear model introduced above applies. 

#### 4. Results
##### Long term impact of biosolids on soil
The normality assumption is validated by the histogram of MWD (Figure 4.1) and normal quantile-quantile plot (Figure 4.2). The histogram shows that the sample distribution of MWD is approximately normal, indicating a normal underlying distribution of MWD. The normal qq plots the sample quantiles from MWD dataset against the theoretical quantiles from the standard normal distribution. The fact that most of the points align well with the straight line suggests normality.

```{r,echo=FALSE}
hist(soil$MWD, xlab="MWD", main = "Histogram of MWD")
```

Figure 4.1 Histogram of MWD

```{r,echo=FALSE}
qqnorm(soil$MWD)
qqline(soil$MWD)
```

Figure 4.2 Normal quantile-quantile plot of MWD

We used ANOVA to compare models with different explanatory variables:  
* A p-value of 0.006 leads to the rejection of the null hypothesis at 1% significance level. There is very strong evidence that a model including treatment effect is very different from that without treatment effect.  
* A p-value less than 0.001 leads to the rejection of the null hypothesis even at 0.1% significance level. There is very strong evidence that a model including date effect is very different from that without date effect.  
* A p-value of 0.008 leads to the rejection of the null hypothesis at 1% significance level. There is very strong evidence that a model including interaction effect of treatment and date is very different from that without interaction effect.

Two treatment types and four sampling dates give us eight combinations. The intercept estimate 0.88 is the mean MWD of control group in April (baseline group). From April to June, the mean MWD of control group increase 0.838. This is the most dramatic increase in MWD among the four sampling dates as Figure 4.3 shows. Moreover, all the estimates for the main effects of dates are positive. It means that MWD always increases as sampling dates change compared to April. As for the main effect of biosolids on soil, mean MWD in biosolids group increases 0.172 compared to the control group in April.

The interaction effect of treatment and sampling date account for the extra change in MWD that is not explained by adding up the main effect of treatment and date when compared with the baseline group. For example, the mean MWD for biosolids group in October is 1.668, which is 0.788 units larger than that for baseline group. There is an increase of 0.376 in mean MWD from April to October. There is another increase of 0.172 in mean MWD from control group to biosolids group. But they do not add up to 0.788. The rest of the difference between the two groups is explained by the interaction effect between biosolids and October (0.24). Visually from the interaction plot (Figure 4.3), we can see that the change in mean MWD over the four sampling dates is quite different for the two treatment groups. Especially from June to August, the mean MWD increase for biosolids group, while the mean MWD decreases for control group. This corroborates the significance of the interaction effect between treatment and dates.  

```{r,echo=FALSE}
Date.f = factor(soil$Date,levels(soil$Date)[c(1, 3, 2, 4)])
ggplot(soil, aes(x = Date.f, y = MWD, group = Treatment, colour = Treatment)) +
  stat_summary(fun.y="mean", geom = "line") +
  labs(x = "Date", title = "Change in MWD over 4 Sampling Dates")
```

Figure 4.3 Change in mean of MWD over the four sampling dates

##### Long term impact of biosolids on plant cover
We check the following assumptions for linear model:  
* Normality assumption is checked by histogram of cover value (Figure 4.4). The distribution of cover values of POPR does not look normal. This is potentially due to the fact that cover value is not normally distributed in nature. Since averaging and using linear regression is our second best solution to deal with discrete cover value, we assume that this violation of normality would not create significant bias on results.  
* Common variance assumption is achieved by a weaker condition that we have equal number of observations (n=4) under biosolids and control.  
* Independence is automatically achieved by averaging dependent observations. Since we only have one observation from each experimental unit (half of block), the observations satisfy independence assumption.

```{r,echo=FALSE}
hist(pc.subset$Cover.value, main="Histogram of cover value for POPR")
```

Figure 4.4 Histogram of cover value for species POPR 

After fitting a regression model with averaged cover value per block as response variable and treatment factor as explanatory variable, we obtain a p-value of 0.00984 for biosolids treatment. With a p-value smaller than 0.01, we find strong evidence to reject the null hypothesis which biosolids and control produce indifferent cover values. The estimated coefficient for biosolids treatment is 25.738. It means that when changing from control to biosolids, we would expect an estimated increase of 25.738 in cover value of POPR.

##### Correlation between MWD and cover value
Since there are unequal number of observations of MWD and cover value, we test correlation using the eight means from the eight combinations of treatment and block for both MWD and cover value. The estimated correlation is 0.645 which indicates moderately strong positive relationship between MWD and cover value. It means that MWD and cover value vary in the same direction and such dependency relationship is moderately strong.

#### 5. Conclusions
There is strong evidence that biosolids application and sampling dates have significant impact on improving soil productivity. They affect MWD interactively instead of individually. There is also strong evidence that biosolids application increases the abundance of certain plant species. There is a moderate positive correlation between MWD and cover value. Better soil quality stimulates more growth of the plants.     

Including treatment, block and date, the models mentioned in Method section are relatively simple but informative about the significance of biosolids application on MWD. A more complicated model would incorporate the nested design in order to capture how transects might affect the conclusion. As for plant species investigation, ordinal regression is more appropriate because cover value is a discrete variable. In terms of the experimental design, we recommend that researchers increase the number of blocks instead of transects in order to increase the number of independent observations. Moreover, soil/plant samples could be taken randomly in blocks instead of in fixed transects. Because proper randomization can balance potentially important explanatory variables at the design stage.

#### 6. Appendix
[link to our Github repository](https://github.com/fjhohura/Stat550)



