---
title: "INITIAL SOIL SCRIPT"
author: "Yidie"
date: "2/19/2017"
output: html_document
---

## LONG TERM IMPACTS OF BIOSOLIDS ON SOILS

### Explore the data
```{r,warning=FALSE,message=FALSE,results='hide'}
library("ggplot2")
library(lme4)
library(reshape2)
library(MASS)
library(dplyr)
```

```{r}
soil<-read.table("MWD.csv", sep = ",", header = T) 
head(soil)
str(soil)
with(soil,table(Treatment, Date))
group_by(soil,Date,Treatment) %>%
  summarise(mean(MWD))
```

### Looking and understanding the data
```{r,echo=TRUE}
hist(soil$MWD, xlab="MWD", main = "Histogram of MWD")

ggplot(aes(x=soil$Treatment,y=soil$MWD), data=soil) + geom_boxplot() + labs(y="MWD", x="Treatment", main="Boxplot of MWD for the Two Treatment Groups")
tapply(soil$MWD, soil$Treatment, mean)
tapply(soil$MWD, soil$Treatment, sd)

Transect.f<-as.factor(soil$Transect)
plot.design(MWD~Treatment+Block+Date+Transect.f, data = soil, xlab="Treatment", ylab="MWD")

ggplot(soil, aes(x = Date, y = MWD, group = Treatment, colour = Treatment)) +
stat_summary(fun.y="mean", geom = "line") +
labs(x = "Date", title = "Change in MWD over 4 Sampling Dates")
```

### Analysis

A simple model to start with:  
- Randomized block design  
- Ignore the transects and repeated measurements for now  
- Treat the data collected from the transects as random samples  
- Assume no block interaction effects

Looking at the data in terms of treatment and block:
```{r,echo=TRUE}
interaction.plot(soil$Treatment, soil$Block, soil$MWD, xlab="Treatment", ylab="MWD", main="Change in MWD over Treatments by Block")
```

```{r,echo=TRUE}
boxcox(MWD~Block+Treatment, data=soil) # To see if transformation on y is needed
soil.rbd<-aov(MWD~Block+Treatment, soil)
summary(soil.rbd)
plot(soil.rbd)
```

Adding Date to the model:
```{r,echo=TRUE}
boxcox(MWD~Block+Treatment*Date, data=soil)
soil.rbd2<-aov(sqrt(MWD)~Block+Treatment*Date, soil)
summary(soil.rbd2)
plot(soil.rbd2)
```

Mixed-effects models (with and without nested factors):
```{r}
lmer1<-lmer(MWD~Block*Date*Treatment+(1|Transect.f), soil)
summary(lmer1)

lmer2<-lmer(MWD~Block*Date*Treatment+(1|Treatment:Transect.f), soil)
summary(lmer2)

lmer3<-lmer(MWD~Block*Date*Treatment+(Date|Treatment:Transect.f), soil)
summary(lmer3)

anova(lmer1,lmer3)
anova(lmer2,lmer3)
```





