---
title: "INITIAL SOIL SCRIPT"
author: "Yidie Feng & Qiaoyue Tang"
date: "2/19/2017"
output: md_document
---

## LONG TERM IMPACTS OF BIOSOLIDS ON MWD

### Explore the data
```{r,warning=FALSE,message=FALSE,results='hide'}
library("ggplot2")
library(lme4)
library(reshape2)
library(MASS)
library(dplyr)
library(base)
```

```{r}
soil<-read.table("MWD.csv", sep = ",", header = T) 
head(soil)
str(soil)
with(soil,table(Treatment, Date))
group_by(soil,Date,Treatment) %>%
  summarise(mean(MWD))
```

### Looking and understanding the data

> S550: Great exploratory analysis. But what is the motivation behind the qqnorm plot? It might not be required for the model we are using. Also, what information does the acf() plot give us? What do you want to show by acf(soil)?

> S550: Some comments above chunks of code would be useful, to state what some of the lesser-known functions do.

> S550: What you get by using this plot design?

```{r,echo=TRUE}
hist(soil$MWD, xlab="MWD", main = "Histogram of MWD")

ggplot(aes(MWD, fill = Treatment, alpha = 0.4), data=soil) + geom_density() + facet_wrap(~Date)
ggplot(aes(y=MWD, x=Treatment, fill = Treatment, alpha = 0.4), data=soil) + geom_boxplot() + geom_point() + facet_wrap(~Date)
tapply(soil$MWD, soil$Treatment, mean)
tapply(soil$MWD, soil$Treatment, sd)

qqnorm(soil$MWD)
qqline(soil$MWD)

acf(soil)

Transect.f<-as.factor(soil$Transect)
plot.design(MWD~Treatment+Block+Date+Transect.f, data = soil, xlab="Treatment", ylab="MWD")

Date.f = factor(soil$Date,levels(soil$Date)[c(1, 3, 2, 4)])
ggplot(soil, aes(x = Date.f, y = MWD, group = Treatment, colour = Treatment)) +
  stat_summary(fun.y="mean", geom = "line") +
  labs(x = "Date", title = "Change in MWD over 4 Sampling Dates")

interaction.plot(soil$Block,soil$Treatment,soil$MWD)
```

### Analysis

A simple model to start with:  
- Randomized complete block design (but block effect as random)  
- Ignore the transects and repeated measurements for now  
- Treat the transects as pseudo-replication

```{r,echo=T}
soil.rcb1<-lmer(MWD~Treatment+(1|Block), soil)  
summary(soil.rcb1)
plot(soil.rcb1)
soil.rcb2<-lmer(MWD~(1|Block), soil)
summary(soil.rcb2)
plot(soil.rcb2)
anova(soil.rcb1,soil.rcb2)
```

Add Date to the model
```{r}
soil.date<-lmer(MWD~Treatment+Date+(1|Block), soil)
summary(soil.date)
plot(soil.date)
anova(soil.rcb1,soil.date)
```

Add interaction effect
```{r,echo=T}
soil.mult<-lmer(MWD~Treatment*Date+(1|Block), soil)
summary(soil.mult)
plot(soil.mult)
anova(soil.date,soil.mult)
```



## LONG TERM IMPACTS OF BIOSOLIDS ON PLANT COVER

### Data Exploration
```{r,echo=TRUE}
pc <- read.csv("plant_cover.csv", header = T)
str(pc)
pc$Block <- as.factor(pc$Block)

# for whole dataset
group_by(pc,Block,Treatment) %>%
  summarise(mean(Cover.value), sd(Cover.value))

table_sum <- tapply(pc$Cover.value,pc$Treatment, table)
table_sum
barplot(table_sum$Biosolids, space = 0, xlab = "Cover Values", ylab = "Frenquency", main = "Counts of POPR with Biosolids Application", col = "mistyrose")
barplot(table_sum$Control, space = 0, xlab = "Cover Values", ylab = "Frenquency", main = "Counts of POPR with Control", col = "mistyrose")

ggplot(aes(x = Block, y = Cover.value, group = Treatment, colour = Treatment), data = pc)+stat_summary(fun.y="mean", geom = "line")+labs(x = "Block", y = "Plant Cover Value", title = "Change in Cover Value over Different Blocks")

# for subset POPR
pc.subset <- subset(pc, Species == "POPR")
table_sum_POPR <- tapply(pc.subset$Cover.value,pc.subset$Treatment, table)
table_sum_POPR
ggplot(aes(x = Block, y = Cover.value, group = Treatment, colour = Treatment), data = pc.subset)+stat_summary(fun.y="mean", geom = "line")+labs(x = "Block", y = "Plant Cover Value", title = "Change in Cover Value of POPR over Different Blocks")

```

### Diagnostics: checking assumptions of linear models
```{r,echo=TRUE}
plot(pc.subset$Treatment, pc.subset$Cover.value)
# similar variability of cover value within two levels of treatment although strange behaviour of Control group

qqnorm(pc.subset$Cover.value)
# strange behaviour due to the fact that cover value is discrete
hist(pc.subset$Cover.value)
# not quite normal, heavy right tail

pc_sub <- pc.subset[ ,c(3,4,5,9)]
acf(pc_sub)
# no multicolinearity within explnatory variables

# Residuals are checked in the next section after imposing linear model.
```

> S550: Here both cover value and cover class are categorical so what does linear regression and diagnostics of the linear regression actually give? How can we check multicolinearity using acf()?



### Analysis
```{r}
pc$Block <- as.factor(pc$Block)
pc.subset <- subset(pc, Species == "POPR")
pc.subset <- pc.subset[ ,c(3,4,9)]

by_blockTrt <- group_by(pc.subset, Block, Treatment)
dat.avg<-summarise(by_blockTrt, y.avg=sum(Cover.value)/50)
add2control <- c("2","Control",0)
dat.avg[8, ] <- add2control
dat.avg

model_avg <- lm(as.numeric(y.avg) ~ Treatment, data = dat.avg)
summary(model_avg)
# MS(Treatment)/MS(Treatment:Block) = F statistic from lm output of adding-zero approach
```

### Correlation between Plant Cover and MWD
```{r}
a<-group_by(soil,Block,Treatment) %>%
  summarise(mean(MWD))
b<-a$`mean(MWD)`
mean.MWD<-c(b[1], b[2], b[3], b[5], b[6], b[7], b[8], b[4])
mean.cv<-as.numeric(as.character(dat.avg$y.avg))
cor(mean.MWD, mean.cv)
```


### Appendix (model history)

#### Part 1

> S550: Good idea to build up the model in this way. But we can take out all the preliminary models when we write the final report.

A simple model to start with:  
- Randomized block design  
- Ignore the transects and repeated measurements for now  
- Treat the data collected from the transects as random samples  
- Assume no block interaction effects

Looking at the data in terms of treatment and block:

interaction.plot(soil$Treatment, soil$Block, soil$MWD, xlab="Treatment", ylab="MWD", main="Change in MWD over Treatments by Block")



boxcox(MWD~Block+Treatment, data=soil) # To see if transformation on y is needed
soil.rbd<-aov(MWD~Block+Treatment, soil)
summary(soil.rbd)
plot(soil.rbd)


Adding Date to the model:

> S550:  Why use boxcox? Is the transformation completely necessary? How much of a difference does it make in the fit? It makes the results less intuitive, so perhaps it's best to leave it out unless the linearity assumption is strongly violated.

> S550: Why use sqrt(MWD) in latter part but not in the upper part?


boxcox(MWD~Block+Treatment*Date, data=soil)
soil.rbd2<-aov(sqrt(MWD)~Block+Treatment*Date, soil)
summary(soil.rbd2)
plot(soil.rbd2)


Mixed-effects model:

Transect.f<-as.factor(soil$Transect)
lmer1.1<-lmer(MWD~Block*Date*Treatment+(1|Transect.f), soil)
summary(lmer1.1)

lmer1.2<-lmer(MWD~Block*Date+(1|Transect.f), soil)
anova(lmer1.1,lmer1.2)

> S550: The three-way intercation probably won't be necessary when choosing the final model. How sure are we that even the two-way interaction is necessary? Perhaps we could just do an additive model?

Mixed-effects model with nested factor:

lmer2.1<-lmer(MWD~Block*Date*Treatment+(1|Treatment:Transect.f), soil)
summary(lmer2.1)

lmer2.2<-lmer(MWD~Block*Date+(1|Treatment:Transect.f), soil)
anova(lmer2.1,lmer2.2)


Mixed-effects model with nested factor and repeated measurements:  
(I am not sure if I formulate the model I want with correct R syntax. I am thinking about fixed block effect, fixed treatment effect, random transect effect, transect factor nested within treatment factor, transect was measured repeatedly over date.)

lmer3.1<-lmer(MWD~Block*Date*Treatment+(Date|Treatment:Transect.f), soil)
summary(lmer3.1)

lmer3.2<-lmer(MWD~Block*Date+(Date|Treatment:Transect.f), soil)
anova(lmer3.1,lmer3.2)


#### Part 2
1) Model 1: all explanatory variables are fixed categorical variables

model1 <- aov(Cover.value ~ Block.f * Treatment * Transect.f)
summary(model1)
plot(resid(model1))
summary(lm(Cover.value ~ Treatment*Block.f*Transect.f))

Comment: Treatment effect seems to be significant. Sum of squares for residuals is very large, indicating a lot of variation is unexplained and the significance of treatment might not be true. No obvious pattern from residual plot.

2) Model 2: Transect is now a random effect, Block and Treatment are fixed effects

model2.1 <- lmer(Cover.value ~ Treatment+Block.f+(1|Transect.f))
summary(model2.1)
model2.2 <- lmer(Cover.value ~ 1+Block+(1|Transect.f))
anova(model2.2, model2.1)


> S550: Not sure if the response variable is categorical then usual linear mixed effect model will be worthy or not? Why you use Block.f for the model2.1 and Block for the model2.2? Also need to clarify about random intercept and random slope model.

Comment: Treatment effect is still significant at the same significance level. From the lm output, A small proportion of randomness is explained by Transect effects but variance of residuals is still quite large. The residuals seem to be centered at 0 from the summary of lm output: scaled residuals.

Question: I did not use aov() because I am not sure how to read summary of aov output properly when there are three explanatory variables. It contains strange behaviours. Not sure if aov output would be more useful here.

3) Model 3: Mixed effects models with nested structure

model3.1 <- lmer(Cover.value ~ Treatment+Block.f+(1|Treatment:Transect.f))
summary(model3.1)
model3.2 <- lmer(Cover.value ~ 1+Block.f+(1|Treatment:Transect.f))
anova(model3.1, model3.2)

Comment: Treatment is still a significant factor in this mixed effect nested model. Same as before, residual variance is still large.

4) Quick comment on Ordinal regression model to incorporate that Cover.value is a discrete dependent variable

Cover.value.f <- as.factor(Cover.value)
model4.1 <- polr(Cover.value.f ~ Treatment+Block.f+Transect.f, Hess = T)
summary(model4.1)
model4.2 <- polr(Cover.value.f ~ 1+Block.f+Transect.f, Hess = T)
anova(model4.1, model4.2)

I tried to include random effects in ordinal regression model but polr() does not recognize the same syntax as above. In addition I get confused when comparing two ordinal regression models as anova() shows a p-value of 0. I think we should discuss whether it is enough to assume that Cover.value is contineous before exploring more about ordinal regression?

detach(pc)

```{r}
# pc.subset <- subset(pc, Species == "POPR")
# subset.cv <- pc.subset[ ,c(3,4,9)]
# 
# Bio_B1_original1 <- 
#   filter(subset.cv, Block == 1 & Treatment == "Biosolids") %>%
#   select(Block)
# zero_B <- list(rep(1, 19))
# a<-rbind(zero_B, Bio_B1_original1)
# 
# Bio_B1_original2 <- 
#   filter(subset.cv, Block == 1 & Treatment == "Biosolids") %>%
#   select(Treatment)
# zero_T <- list(rep("Biosolids", 19))
# b<-rbind(zero_T, Bio_B1_original2)
# 
# Bio_B1_original3 <- 
#   filter(subset.cv, Block == 1 & Treatment == "Biosolids") %>%
#   select(Cover.value)
# zero_C <- list(rep(0, 19))
# c<-rbind(zero_C, Bio_B1_original3)
# 
# Bio_B1 <- cbind(a,b,c)
# 
# #----
# 
# Bio_B2_original1 <- 
#   filter(subset.cv, Block == 2 & Treatment == "Biosolids") %>%
#   select(Block)
# zero_B <- list(rep(2, 36))
# a<-rbind(zero_B, Bio_B2_original1)
# 
# Bio_B2_original2 <- 
#   filter(subset.cv, Block == 2 & Treatment == "Biosolids") %>%
#   select(Treatment)
# zero_T <- list(rep("Biosolids", 36))
# b<-rbind(zero_T, Bio_B2_original2)
# 
# Bio_B2_original3 <- 
#   filter(subset.cv, Block == 2 & Treatment == "Biosolids") %>%
#   select(Cover.value)
# zero_C <- list(rep(0, 36))
# c<-rbind(zero_C, Bio_B2_original3)
# 
# Bio_B2 <- cbind(a,b,c)
# 
# #------
# 
# Bio_B3_original1 <- 
#   filter(subset.cv, Block == 3 & Treatment == "Biosolids") %>%
#   select(Block)
# zero_B <- list(rep(3, 21))
# a<-rbind(zero_B, Bio_B3_original1)
# 
# Bio_B3_original2 <- 
#   filter(subset.cv, Block == 3 & Treatment == "Biosolids") %>%
#   select(Treatment)
# zero_T <- list(rep("Biosolids", 21))
# b<-rbind(zero_T, Bio_B3_original2)
# 
# Bio_B3_original3 <- 
#   filter(subset.cv, Block == 3 & Treatment == "Biosolids") %>%
#   select(Cover.value)
# zero_C <- list(rep(0, 21))
# c<-rbind(zero_C, Bio_B3_original3)
# 
# Bio_B3 <- cbind(a,b,c)
# 
# #-----------------------
# Bio_B4_original1 <- 
#   filter(subset.cv, Block == 4 & Treatment == "Biosolids") %>%
#   select(Block)
# zero_B <- list(rep(4, 9))
# a<-rbind(zero_B, Bio_B4_original1)
# 
# Bio_B4_original2 <- 
#   filter(subset.cv, Block == 4 & Treatment == "Biosolids") %>%
#   select(Treatment)
# zero_T <- list(rep("Biosolids", 9))
# b<-rbind(zero_T, Bio_B4_original2)
# 
# Bio_B4_original3 <- 
#   filter(subset.cv, Block == 4 & Treatment == "Biosolids") %>%
#   select(Cover.value)
# zero_C <- list(rep(0, 9))
# c<-rbind(zero_C, Bio_B4_original3)
# 
# Bio_B4 <- cbind(a,b,c)
# 
# #-----------------------
# 
# Con_B1_original1 <- 
#   filter(subset.cv, Block == 1 & Treatment == "Control") %>%
#   select(Block)
# zero_B <- list(rep(1, 47))
# a<-rbind(zero_B, Con_B1_original1)
# 
# Con_B1_original2 <- 
#   filter(subset.cv, Block == 1 & Treatment == "Control") %>%
#   select(Treatment)
# zero_T <- list(rep("Control", 47))
# b<-rbind(zero_T, Con_B1_original2)
# 
# Con_B1_original3 <- 
#   filter(subset.cv, Block == 1 & Treatment == "Control") %>%
#   select(Cover.value)
# zero_C <- list(rep(0, 47))
# c<-rbind(zero_C, Con_B1_original3)
# 
# Con_B1 <- cbind(a,b,c)
# 
# #----
# 
# a <- matrix(2, 50, 1)
# 
# b <- matrix("Control", 50,1)
# 
# c <- matrix(0, 50,1)
# 
# Con_B2<-data.frame(cbind(a,b,c))
# colnames(Con_B2)[1]<-"Block"
# colnames(Con_B2)[2]<-"Treatment"
# colnames(Con_B2)[3]<-"Cover.value"
# 
# 
# 
# #-------
# 
# Con_B3_original1 <- 
#   filter(subset.cv, Block == 3 & Treatment == "Control") %>%
#   select(Block)
# zero_B <- list(rep(3, 49))
# a<-rbind(zero_B, Con_B3_original1)
# 
# Con_B3_original2 <- 
#   filter(subset.cv, Block == 3 & Treatment == "Control") %>%
#   select(Treatment)
# zero_T <- list(rep("Control", 49))
# b<-rbind(zero_T, Con_B3_original2)
# 
# Con_B3_original3 <- 
#   filter(subset.cv, Block == 3 & Treatment == "Control") %>%
#   select(Cover.value)
# zero_C <- list(rep(0, 49))
# c<-rbind(zero_C, Con_B3_original3)
# 
# Con_B3 <- cbind(a,b,c)
# 
# 
# #-------------------
# 
# 
# Con_B4_original1 <- 
#   filter(subset.cv, Block == 4 & Treatment == "Control") %>%
#   select(Block)
# zero_B <- list(rep(4, 49))
# a<-rbind(zero_B, Con_B4_original1)
# 
# Con_B4_original2 <- 
#   filter(subset.cv, Block == 4 & Treatment == "Control") %>%
#   select(Treatment)
# zero_T <- list(rep("Control", 49))
# b<-rbind(zero_T, Con_B4_original2)
# 
# Con_B4_original3 <- 
#   filter(subset.cv, Block == 4 & Treatment == "Control") %>%
#   select(Cover.value)
# zero_C <- list(rep(0, 49))
# c<-rbind(zero_C, Con_B4_original3)
# 
# Con_B4 <- cbind(a,b,c)
# 
# #----------------------
# dat <- rbind(Bio_B1, Bio_B2, Bio_B3, Bio_B4, Con_B1, Con_B2,Con_B3,Con_B4)
# 
# model_zero <- aov(as.numeric(Cover.value) ~ Treatment+Error(Treatment/Block), data = dat)
# summary(model_zero)
# 
# detach(pc)
# ```
# 
# ```{r}
# pc.corr <- melt(tapply(pc$Cover.value, pc$Block, mean))
# soil.june <- soil %>%
#   filter(Date == "June")
# soil.corr <- melt(tapply(soil.june$MWD, soil.june$Block, mean))
# cor(pc.corr[2], soil.corr[2])
```

MWD and Plant Cover Value seem to have a strong negative correlation which is strange because we expect to see at least a positive correlation. The problem is that the two variable, MWD and Cover Value, do not have the same length. MWD is measured in 4 months and is calculated as an average of 3 transects. Cover Value is measured only once in June and is calculated as an average of 5 transects. In the code above, I simply take all average within blocks (4 measurements for each MWD and plant cover data) and take only June measurements from MWD data. We should fix this issue soon in next steps.

> S550: why use melt function here? only tapply gives the same result. But not sur about the process!

> S550: We might have to do a Spearman correlation (it's an option inside the cor() function), because one of the variables is ordinal.
